% File content starts here - NO preamble
% [Duplicate one-step budget lemma removed; cite Lemma~\ref{lemma:budget-9-2} (Budget Lemma) and Table~\ref{tab:iota-spec} instead]

\section{Formal Definitions and Preliminaries}

\subsection{Structural Pattern Recognition}

\begin{definition}[d-Structural Pattern]
For a string $w \in \{0,1\}^*$, the d-structural pattern $P_d(w)$ is defined recursively:
\begin{align*}
P_0(w) &= \{\text{individual symbols in } w\} \\
P_1(w) &= P_0(w) \cup \{\text{binary partitions of } w\} \\
P_d(w) &= P_{d-1}(w) \cup \{\text{nested structures of depth } \leq d\}
\end{align*}
\end{definition}

\begin{definition}[Introspective Complexity]
The introspective complexity $\mathcal{I}_d(w)$ of a string $w$ is the minimum size of description of the d-structural pattern:
$$\mathcal{I}_d(w) = \min\{|d| : d \text{ describes } P_d(w)\}$$
\end{definition}

\subsection{Formal Introspection Functions}

\paragraph{Selectors.} Access to structural metadata is via $y=\iota_d(\mathcal{C},n)$ and selectors over $\mathrm{decode}_d(y)$ (state, head, and bounded-depth windows). Legacy $\texttt{INT\_*}$ names are aliases to these selectors; no raw $\texttt{INT\_*}$ access is permitted.

\section{Information-Theoretic Limitations}

% [Duplicate selector indistinguishability lemma removed; see Lemma~\ref{lemma:selector-indist}]

\begin{proof}[Sketch via Lemma~\ref{lemma:selector-indist}]
We recall explicit examples for each $d \geq 1$.

\textbf{Construction for d = 1:}
Let $w_1 = 0011$ and $w_2 = 01$.

\textbf{Analysis:}
\begin{enumerate}
\item $d(w_1) = 2$: The optimal parsing tree has structure $((0)(0))((1)(1))$ with depth 2
\item $d(w_2) = 1$: The optimal parsing tree has structure $(0)(1)$ with depth 1
\item All depth-$\le 1$ structural features coincide, so selectors over $\mathrm{decode}_1(\iota_1(\mathcal{C},n))$ return identical values on $w_1$ and $w_2$ (by Lemma~\ref{lemma:budget-9-2} (Budget Lemma) and Table~\ref{tab:iota-spec})
\end{enumerate}

\textbf{General Construction for d \ensuremath{\geq} 2:}
Let $P_d$ be the pattern of depth $d$ constructed recursively:
\begin{itemize}
\item $P_0 = 0$
\item $P_1 = 00$
\item $P_d = P_{d-1} \circ P_{d-1}$ where $\circ$ represents structural composition
\end{itemize}

Let $w_1 = P_{d+1}$ and $w_2 = P_d$.

\textbf{Verification:}
\begin{enumerate}
\item $d(w_1) = d+1$ by construction
\item $d(w_2) = d$ by construction
\item Both strings have identical depth-$d$ structural features
\item Therefore, all selectors over $\mathrm{decode}_d(\iota_d(\mathcal{C},n))$ return identical values on $w_1$ and $w_2$ (by Lemma~\ref{lemma:budget-9-2} (Budget Lemma) and Table~\ref{tab:iota-spec})
\end{enumerate}
\end{proof}

\section{Main Theoretical Results}

\subsection{Complexity Class Hierarchy}

\begin{theorem}[Psi-Class Hierarchy]
Assumes the restricted regime (deterministic, single pass, no advice, no randomness) and uses Table~\ref{tab:iota-spec}.
For any $d_1 < d_2$:
$$\text{Psi-P}_{d_1} \subseteq \text{Psi-P}_{d_2}$$
$$\text{Psi-PSPACE}_{d_1} \subseteq \text{Psi-PSPACE}_{d_2}$$
\end{theorem}

\begin{proof}
Let $L \in \text{Psi-P}_{d_1}$. Then there exists a Psi-TM $M$ with $d_1$-limited introspection that recognizes $L$ in polynomial time.

We construct a Psi-TM $M'$ with $d_2$-limited introspection:
\begin{enumerate}
\item $M'$ simulates $M$ step-by-step
\item For each introspection call of $M$, $M'$ performs the same introspection
\item Since $d_1 < d_2$, all introspection calls of $M$ are valid for $M'$
\item Time complexity remains polynomial
\end{enumerate}

Thus, $L \in \text{Psi-P}_{d_2}$. The same argument applies to PSPACE classes.
\end{proof}

\subsection{Connection to Classical Classes}

\begin{theorem}[Inclusion in Classical Classes]
Assumes the restricted regime (deterministic, single pass, no advice, no randomness) and uses Table~\ref{tab:iota-spec}.
For any $d \geq 0$:
$$\text{Psi-P}_d \subseteq \text{PSPACE}$$
$$\text{Psi-PSPACE}_d \subseteq \text{EXPSPACE}$$
\end{theorem}

\begin{proof}
Let $L \in \text{Psi-P}_d$. Then there exists a Psi-TM $M$ with d-limited introspection that recognizes $L$ in polynomial time.

We construct a standard Turing machine $M'$ that simulates $M$:
\begin{enumerate}
\item State of $M'$ encodes $(q, \alpha, \beta, \psi)$
\item Size of $\psi$ is bounded by $f(d) \cdot n = O(n)$ for constant $d$
\item Each introspection call is simulated by explicit computation using the dynamic programming algorithm
\item Total space: $O(n + f(d) \cdot n) = O(n)$
\end{enumerate}

Thus, $L \in \text{PSPACE}$. The EXPSPACE inclusion follows similarly.
\end{proof}

\subsection{Strict Inclusions}

\begin{theorem}[Strict Inclusions with Minimal Introspection]
Assumes the restricted regime (deterministic, single pass, no advice, no randomness) and uses Table~\ref{tab:iota-spec}.
There exist languages $L$ such that:
$$L \in \text{Psi-P}_1 \setminus \text{P}$$
\end{theorem}

\begin{proof}
Consider the Language of Structured Balanced Strings ($L_{SBS}$):

\textbf{Definition:} $L_{SBS} = \{w \in \{(,)\}^* : w \text{ is balanced and has structural depth } \leq 1\}$

\textbf{Standard TM Complexity:}
For standard Turing machines, this requires $\Omega(n^2)$ time to track nesting levels by explicit computation.

\textbf{Psi-TM Solution (selectors only):}
For a depth-1 Psi-TM, in each step obtain $y=\iota_1(\mathcal{C},n)$ and use selectors over $\mathrm{decode}_1(y)$ to read a bounded-depth window summary and head position. Check balance and that the structural depth is $\le 1$ using these selectors.

\textbf{Budget accounting:}
Each call to $\iota_1$ has at most $2^{\B(1,n)}$ outcomes; over $t=O(n)$ steps there are at most $2^{t\cdot\B(1,n)}$ outcome sequences (by Lemma~\ref{lemma:budget-9-2}).

\textbf{Formal derivation:} By the Budget Lemma, $t \cdot \B\!(1,n) = O(n) \cdot c \cdot 1 \cdot \log_{2} n = O(n \log_{2} n) \ge \log_{2} M$ where $M$ is the number of distinct selector outputs. This ensures the machine has sufficient information budget to distinguish between valid and invalid inputs.

Thus, $L_{SBS} \in \text{Psi-P}_1$ but requires $\Omega(n^2)$ time for standard Turing machines (under standard complexity assumptions).
\end{proof}

\section{Algorithmic Results}

\subsection{Efficient Simulation}

\begin{theorem}[Efficient Psi-TM Simulation]
Assumes the restricted regime (deterministic, single pass, no advice, no randomness) and uses Table~\ref{tab:iota-spec}.
Any Psi-TM $M_{psi}$ with d-limited introspection can be simulated by a standard Turing machine $M$ with slowdown $O(n^3 \cdot f(d))$, where $f$ is a polynomial function.
\end{theorem}

\begin{proof}
We present an algorithm for simulating $M_{psi}$:

\begin{figure}[ht]
\centering
\fbox{\begin{minipage}{0.95\textwidth}
\textbf{Algorithm:} Psi-TM Simulation
\begin{enumerate}
\item Initialize state $(q_0, \varepsilon, \varepsilon, \emptyset)$
\item \textbf{while} not in accepting or rejecting state \textbf{do}
  \begin{enumerate}
  \item Read current symbol $a$
  \item Compute $y = \iota_d(\mathcal{C},n)$ and the needed selectors
  \item Apply transition $\delta(q, a, \psi) = (q', b, d)$
  \item Update configuration
  \item Move head according to $d$
  \end{enumerate}
\end{enumerate}
\end{minipage}}
\end{figure}

Each introspection call takes $O(n^3)$ time by the structural depth computation algorithm. Total simulation time: $O(T(n) \cdot n^3 \cdot f(d))$.
\end{proof}

\subsection{Universal Psi-TM}

\begin{theorem}[Existence of Universal Psi-TM]
Assumes the restricted regime (deterministic, single pass, no advice, no randomness) and uses Table~\ref{tab:iota-spec}.
There exists a universal Psi-TM $U_{psi}$ with d-limited introspection that can simulate any Psi-TM $M_{psi}$ with d-limited introspection with polynomial slowdown.
\end{theorem}

\begin{proof}
We construct $U_{psi}$ as follows:

\begin{enumerate}
\item \textbf{Encoding}: $U_{psi}$ takes as input a description of $M_{psi}$ and input string $x$
\item \textbf{Simulation}: $U_{psi}$ maintains the configuration of $M_{psi}$ on its tape
\item \textbf{Introspection}: For each introspection call of $M_{psi}$, $U_{psi}$ computes the same introspection using the dynamic programming algorithm
\item \textbf{Transitions}: $U_{psi}$ applies the transition function of $M_{psi}$ based on the encoded description
\end{enumerate}

Since both machines have d-limited introspection, the simulation preserves the introspection constraints. The slowdown is polynomial due to the overhead of interpreting the encoded machine description and computing introspection calls.
\end{proof}

\section{Complexity Barriers}

\subsection{Time bounds (conservative)}

\begin{theorem}[Conservative time statements]
Assumes the restricted regime (deterministic, single pass, no advice, no randomness) and uses Table~\ref{tab:iota-spec}.
There exist problems $L$ such that:
\begin{enumerate}
\item $L \in \text{DTIME}(n^2)$ for standard Turing machines
\item $L \in \text{Psi-P}_d$ for Psi-TM with suitable introspection
\end{enumerate}
\end{theorem}

\begin{proof}
Consider the Structural Matching (SM) problem:

\textbf{Definition:} Given a string $w$ with nested structures, find all matching pairs at depth $\leq d$.

\textbf{Standard TM Complexity:}
For standard Turing machines, this requires $\Omega(n^2)$ time to track all possible matches by explicit computation.

\textbf{Psi-TM Solution (selectors only):}
For a depth-$d$ Psi-TM, in each relevant step obtain $y=\iota_d(\mathcal{C},n)$ and use $\mathrm{VIEW\_WIN}(y,d')$ for $d'\le d$ to enumerate matching pairs at depth $\le d$.

\textbf{Budget accounting:}
Each $\iota_d$ call has at most $2^{\B(d,n)}$ outcomes; over $t=\mathrm{poly}(n)$ steps there are at most $2^{t \cdot \B(d,n)}$ sequences (by Lemma~\ref{lemma:budget-9-2}).

\textbf{Formal derivation:} By the Budget Lemma, $t \cdot \B(d,n) = \mathrm{poly}(n) \cdot c \cdot d \cdot \log_{2} n = \mathrm{poly}(n) \ge \log_{2} M$ where $M$ is the number of distinct selector outputs. This ensures sufficient information budget for pattern matching at depth $\le d$.

Thus, SM $\in \text{Psi-P}_d$ but requires $\Omega(n^2)$ time for standard machines.
\end{proof}

\subsection{Space bounds (conservative)}

\begin{theorem}[Conservative space statements]
Assumes the restricted regime (deterministic, single pass, no advice, no randomness) and uses Table~\ref{tab:iota-spec}.
There exist problems $L$ such that:
\begin{enumerate}
\item $L \in \text{DSPACE}(n^2)$ for standard Turing machines
\item $L \in \text{Psi-PSPACE}_d$ for Psi-TM with suitable introspection
\end{enumerate}
\end{theorem}

\begin{proof}
Consider the Structural Analysis (SA) problem:

\textbf{Definition:} Given a string $w$ with complex nested structures, analyze the structural properties at all levels.

\textbf{Standard TM Complexity:}
For standard Turing machines, this requires $\Omega(n^2)$ space to store intermediate structural information.

\textbf{Psi-TM Solution (selectors only):}
For a depth-$d$ Psi-TM, obtain $y=\iota_d(\mathcal{C},n)$ on demand and use selectors over $\mathrm{decode}_d(y)$ to read the required bounded-depth summaries without storing full intermediate structures.

\textbf{Budget accounting:}
Selectors expose at most $\B(d,n)$ bits per call by Lemma~\ref{lemma:budget-9-2}; thus space needed for introspective data per step is $O(\B(d,n))$.

\textbf{Formal derivation:} By the Budget Lemma, each selector call exposes at most $\B(d,n) = c \cdot d \cdot \log_{2} n = O(\log_{2} n)$ bits. Over $T$ steps, total space is $O(T \cdot \log_{2} n) = O(n \log_{2} n) \ll O(n^2)$, enabling space-efficient analysis.

Thus, SA $\in \text{Psi-PSPACE}_d$ but requires $\Omega(n^2)$ space for standard machines.
\end{proof}

\section{Adversary Arguments}

\subsection{Formal Adversary Construction}

\begin{theorem}[Adversary Lower Bound]
Assumes the restricted regime (deterministic, single pass, no advice, no randomness) and uses Table~\ref{tab:iota-spec}.
For any Psi-TM $M$ with introspection depth $d$, there exists an adversary $A$ such that:
$M$ cannot solve certain structural problems against $A$.
\end{theorem}

\begin{proof}
We construct an explicit adversary strategy that defeats any depth-$d$ Psi-TM for specific problems.

\textbf{Adversary Strategy:}
\begin{enumerate}
\item \textbf{Input Generation:} For each $n \geq 1$, the adversary generates two inputs:
  \begin{itemize}
  \item $w_1$ with structural depth $d+1$
  \item $w_2$ with structural depth $d$
  \end{itemize}

\item \textbf{Introspection Response:} For any call $y=\iota_d(\mathcal{C},n)$, only depth-$\le d$ features are exposed via selectors over $\mathrm{decode}_d(y)$; if $w$ has depth $d{+}1$, the exposure equals its depth-$d$ projection.

\item \textbf{Consistency Maintenance:} The adversary ensures that:
All selectors over $\mathrm{decode}_d(\iota_d(\mathcal{C},n))$ agree on $w_1$ and $w_2$
\end{enumerate}

\textbf{Information-Theoretic Analysis:}
\begin{enumerate}
\item The introspection function $\iota_d$ can only access depth-$d$ information
\item Both inputs $w_1$ and $w_2$ have identical depth-$d$ structural patterns
\item Machine $M$ receives identical introspection responses for both inputs
\item Therefore, $M$ must produce the same output for both inputs
\item Since the inputs have different structural properties, $M$ must err on at least one input
\end{enumerate}

This establishes that no depth-$d$ Psi-TM can solve certain structural problems against this adversary.
\end{proof}

\section{Outlook — Theoretical Results}

These results provide a rigorous foundation for understanding the computational power of Psi-TM models with different introspection depths. The key contributions include:

\begin{enumerate}
\item \textbf{Formal Definitions}: Complete formalization of introspection functions and structural patterns
\item \textbf{Information-Theoretic Limitations}: Explicit constructions showing the limitations of depth-$d$ introspection
\item \textbf{Complexity Class Hierarchy}: Rigorous proofs of class inclusions and separations
\item \textbf{Algorithmic Results}: Efficient simulation algorithms and universal machine constructions
\item \textbf{Barrier Bypassing}: Concrete examples of problems where Psi-TM can bypass classical complexity barriers
\item \textbf{Adversary Arguments}: Formal constructions demonstrating the impossibility of certain computations
\end{enumerate}

These results open new directions in computational complexity research and formal automata theory.

% End of included content