% Controlled Relaxations (v0.9.1)
\subsection{Controlled relaxations}
We record four relaxations and their explicit polynomial degradations.

\begin{table}[t]
\centering
\caption{Controlled Relaxations (v0.9.1): Relaxation $\to$ Degradation bound $\to$ Reference}
\begin{tabular}{lll}
\toprule
Relaxation & Degradation bound & Reference \\
\midrule
R1: Public randomness & explicit in entropy budget $H$ (state dependence on $H$) & Lemma~\ref{lem:R1-entropy-budget} \\
R2: Multi-pass ($P$ passes) & $P\cdot(m+B(d,n)) \ge c_0\cdot \log M$ (state origin of $c_0$) & Lemma~\ref{lem:R2-multipass} \\
R3: Advice & survives for $o(n)$ and $O(\log n)$; show explicit dependence on $|adv|$ & Lemma~\ref{lem:R3-advice} \\
R4: Bandwidth tweak $\pm O(\log n)$ & LB stable under small shifts in $B(d,n)$; bound losses & Lemma~\ref{lem:R4-bandwidth-tweak} \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Lemma (R1: entropy budget).}\label{lem:R1-entropy-budget}
Public randomness with entropy budget $H$ perturbs the information budget by at most an additive $H$ term. By Budget accounting and $\Psi$-Fano/\,$\Psi$-fooling arguments, the lower bound degrades polynomially with explicit dependence on $H$ in the state space; no change to asymptotic exponents beyond this parameterization.

\paragraph{Lemma (R2: multi-pass).}\label{lem:R2-multipass}
For $P$ passes, if $P\cdot(m+\B(d,n)) \ge c_0\cdot \log M$, the bound persists with polynomial loss. The constant $c_0$ arises from the query-to-entropy conversion in the Budget Lemma (normalizing per-step payload to $\B(d,n)$ and per-pass overhead $m$) and the base-2 encoding used in transcript counting.

\paragraph{Lemma (R3: advice).}\label{lem:R3-advice}
With advice $|adv| \in \{O(\log n),\, o(n)\}$, the lower bound is stable: the proof augments the information budget by $|adv|$ bits and then applies the $\Psi$-Fooling/\,$\Psi$-Fano tool. Dependence on $|adv|$ is explicit and purely additive inside the budget term.

\paragraph{Lemma (R4: bandwidth tweak).}\label{lem:R4-bandwidth-tweak}
For bandwidth shifts $\B(d,n) \mapsto \B(d,n) \pm O(\log n)$, the transcript and budget inequalities change by polylogarithmic factors. Anti-simulation and budget monotonicity yield that the lower bound is preserved up to polylogarithmic losses with constants inherited from the corresponding tools.