% v0.8.7 — Reproducibility & Artifact Summary (Self-Contained)

\section{Reproducibility \& Artifact Summary (Self-Contained)}
\label{sec:repro}

This section summarizes what is included in the \texttt{v0.8.7} drop and the minimal steps a reader can take to verify correctness \emph{without consulting external material}. All statements use base-2 logarithms (\(\log_{2}\)).

\subsection{Scope and Artifacts}
\begin{itemize}
  \item Formal model, lower-bound tools, and barrier statements with Lean skeletons (headers type-checked; proofs may be partial in the 0.8.6.x series).
  \item Claims index mapping LaTeX labels to Lean IDs (kept in sync by validators).
  \item Embedded figures and representative plots for key results.
  \item Appendix ``Pre-Fortress Firebreak'': stress tests and red-team outcomes.
  \item Deterministic notebooks; all experiments use fixed seed \(1337\).
\end{itemize}

\subsection{Minimal Reproducibility Commands (Optional)}
Environment: Lean~4.22.x + Lake~5.x, Python~3.10--3.12, \TeX~Live~2021+.
\begin{enumerate}
  \item \textbf{Lean build (type-checks skeletons)}\\
  \texttt{lake build}
  \item \textbf{Validators (claims and project checks)}\\
  \texttt{python scripts/validate\_claims.py\ paper/claims.yaml}\\
  \texttt{python scripts/check\_project.py}
  \item \textbf{PDF build (\LaTeX)}\\
  \texttt{latexmk -pdf -halt-on-error -interaction=nonstopmode main.tex}
\end{enumerate}

\subsection{Acceptance Criteria}
The artifact is considered consistent if: (i) \LaTeX\ compiles without errors; (ii) the claims validator reports all mappings consistent; (iii) figures referenced in the text are embedded (no external paths); (iv) determinism is documented (seed~\(1337\)).

\subsection{Guard Rails (Conventions)}
\begin{itemize}
  \item Base-2 logarithms only (\(\log_{2}\)); Lean code uses \texttt{Nat.log2} exclusively.
  \item Deterministic execution in notebooks with seed \(1337\).
\end{itemize}

\paragraph{Artifact Availability.}
Code and data are released under \texttt{CC BY-NC-ND 4.0}. A tagged snapshot (\texttt{v0.8.7}) accompanies this paper (DOI if applicable).

